Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-11-19 doc_arXiv/doc_arXiv.tex
feedback from RC

\author{
    Byron C. Jaeger
    \thanks{Source code available at
\url{https://github.com/bcjaeger/INTERMACS-missing-data}}
   \\
    Department of Biostatistics \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
   \And
    Ryan Cantor, PhD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Venkata Sthanam, MS
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Rongbing Xie, DrPH
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    James K. Kirklin, MD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Ramaraju Rudraraju, PhD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:ramaraju@uab.edu}{\nolinkurl{ramaraju@uab.edu}}} \\
  }

@@

\begin{abstract}
\textbf{Background} Risk prediction models play an important role in
clinical decision making. When developing risk prediction models,
practitioners often impute missing values to the mean. We evaluated the
impact of applying other strategies to impute missing values on the
prognostic accuracy of downstream risk prediction models, i.e., models
fitted to the imputed data. A secondary objective was to compare the
accuracy of imputation methods based on artificially induced missing
values. To complete these objectives, we used data from the Interagency
Registry for Mechanically Assisted Circulatory Support (INTERMACS).
\newline\textbf{Methods and Results} We applied twelve imputation
strategies in combination with two different modeling strategies for
mortality and transplant risk prediction following surgery to receive
mechanical circulatory support. Model performance was evaluated using
Monte-Carlo cross validation and measured based on outcomes 6-months
following surgery using the scaled Brier score, concordance index, and
calibration error. We used Bayesian hierarchical models to compare model
performance. Multiple imputation with random forests emerged as a robust
strategy to impute missing values, increasing model concordance by 0.003
(25th, 75th percentile: 0.008, 0.052) compared with imputation to the
mean for mortality risk prediction using a downstream proportional
hazards model. The posterior probability that single and multiple
imputation using random forests would improve concordance versus mean
imputation was 0.482 and 0.962, respectively.
\newline\textbf{Conclusion} Selecting an optimal strategy to impute
missing values and applying multiple imputation can improve the
prognostic accuracy of downstream risk prediction models.
\end{abstract}

\keywords{
    Missing Data
   \and
    INTERMACS
   \and
    imputation
   \and
    heart failure
   \and
    mortality
   \and
    risk prediction
  }

@@

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
A critical step for developing risk prediction models is handling
missing data, which is common in a ``real world'' database that is not
designed for the rigor of data completeness found in a clinical trial.
While some missing values are related to data entry error, others may
relate to instability of a patient's circulatory status at the time of
implant. For instance, hemodynamic laboratory values are missing for
patients in whom invasive catheterizations were not performed. Patients
without invasive catheterizations may be too ill (unstable) to tolerate
the procedure. Instead, such patients may, after determination of basic
hemodynamics, proceed directly to device placement.

The primary aim of this article is to quantify how much the prognostic
value of a risk prediction model developed from the INTERMACS registry
depends on the strategy that was applied to impute missing data prior to
developing the model. A secondary aim is to measure imputation accuracy
of each strategy by introducing varying levels of artificial missing
data and then imputing it. Because imputation to the mean has been a
standard method for multiple annual summaries of the INTERMACS data, we
measure the potential improvement in prognostic value of a risk
prediction model when other strategies are applied to impute missing
data \emph{instead of imputation to the mean}. The over-arching aim is
to clarify which imputation strategies are most likely to improve risk
prediction (and by extension, quality of care) for patients who receive
MCS devices. This investigation can directly inform future analyses of
INTERMACS data and provide evidence quantifying the benefit of imputing
missing data with sound methodology.

@@

Patient follow-up begins after implantation of a durable, long term MCS
device and continues while the device is in place. Registry endpoints
include death on a device, heart transplantation, or cessation of
support (for recovery and non-recovery reasons). Mortality and
transplant after MCS were the primary outcomes for the current study. As
there were only 310 cessation of support events, we did not analyze this
outcome. For mortality and transplant risk prediction models, we applied
cause-specific analysis to account for competing risks. For example, in
risk prediction models for mortality, patients were censored at time of
loss to follow-up, transplant, or cessation of support, whichever
occurred first. INTERMACS collects pre-implant data on patient
characteristics, medical status, and laboratory values. INTERMACS also
collects follow-up data at regularly scheduled visits and at occurrence
of adverse events such as re-hospitalization. For the current analysis,
all pre-implant variables were considered as potential predictors.

@@

To conduct statistical inference in the context of missing data setting,
analysts often create multiple imputed datasets, replicate an analysis
in each of them, and then pool their results to obtain valid test
statistics for hypothesis testing \cite{rubin2004multiple}. Imputation
strategies that create a single imputed dataset
(\textit{e.g., }imputation to the mean) have been shown to increase type
I errors (\textit{i.e., }rejecting a true null hypothesis) for
inferential statistics by artificially reducing the variance of observed
data and ignoring the uncertainty attributed to missing values
\cite{van2018flexible}. To ensure valid inference, imputation models
should leverage outcome variables to predict missing predictors
\cite{sterne2009multiple}. When very few data are missing, analysts may
apply listwise deletion, \textit{i.e., }removing any observation with at
least one missing value. However, listwise deletion can easily lead to
biased inference \cite{van2020rebutting}.

@@

In the presence of missing data, the goal of supervised statistical
learning is to develop a prediction function for an outcome variable
that accurately generalizes to testing data, which may or may not
contain missing values \cite{hastie2009elements}. Because testing data
may contain missing values, listwise deletion is not a feasible strategy
for statistical learning tasks. In contrast to statistical inference,
strategies that create a single imputed dataset are often used for
statistical learning \cite{kuhn2019feature}. Previous work has
emphasized the importance of imputation strategies with greater
accuracy, suggesting that more accurate imputation strategies lead to
better performance of downstream models (\textit{i.e., }models fitted to
the imputed data) \cite{jerez2010missing}. Others have shown that
imputation to the mean, model-based imputation, and multiple imputation
can provide Bayes-optimal prediction models provided certain assumptions
are true, but these assumptions are difficult to validate in applied
settings \cite{josse2019consistency}.

@@

KNN imputation identifies \(k\) `similar' observations
(\textit{i.e., }neighbors) for each observation with a missing value in
a given variable \cite{chen2000nearest}. A donor value for the current
missing value is generated by sampling or aggregating the values from
the \(k\) nearest neighbors. In the current analysis, we identified 10
nearest neighbors using Gower's distance \cite{gower}. When imputing a
single dataset using KNN, we aggregated values from 10 nearest neighbors
using the median for numeric variables and the mode for categorical
variables. When imputing multiple datasets using KNN, we aggregated
values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5
imputed datasets.

@@

Random forests grow an ensemble of de-correlated decision trees, where
each tree is grown using a bootstrapped replicate of the original
training data
\cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
A particularly helpful feature of random forests is their ability to
estimate testing error by aggregating each decision tree's prediction
error on data outside of their bootstrapped sample
(\textit{i.e., }out-of-bag error). In the current analysis, we conduct
MICE using one random forest to impute each variable, separately. For
each imputed dataset, we allowed random forests to be re-fitted until
out-of-bag error stabilized or a maximum number of iterations was
completed. When imputing a single dataset, we used 250 trees per forest
and a maximum of 10 iterations. When imputing multiple datasets, we used
50 trees per forest and applied PMM using the random forest's predicted
values to impute missing data by sampling one value from a pool of 10
potential donors.

@@

MIA is a technique that uses missing status as a predictor rather than
explicitly imputing missing values
\cite{twala2009empirical, ding2010investigation}. MIA adds another
category to nominal variables: ``Missing''. For numeric variables, MIA
creates two columns: one where missing values are imputed with positive
infinity and the other with negative infinity. Since PH models are not
compatible with infinite values, we only use MIA in boosting models.
When a decision tree uses a finite cut-point to split a given numeric
variable, it will assess the cut-point using both the positive and
negative infinite imputed columns, and utilize whichever column provides
the best split of the current data. This procedure translates to sending
all missing values to the left or to the right when forming two new
nodes of the decision tree, using whichever direction results in a
better split.

@@

Imputation accuracy was computed for each numerical and nominal
variable, separately. Numeric variable imputation accuracy was measured
using a re-scaled mean-squared error:
\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
MSEs are equal, and less than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
imputation accuracy was measured using a re-scaled classification
accuracy:
\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if the classification error of the current
imputation method is less than that of imputation to the mean, equal to
0 if the two classification errors are equal, and less than 0 of the
classification error of the current imputation method is worse than
imputation to the mean. These numeric and nominal scores are analogous
to the more well known \(R^2\) and Kappa statistics, respectively, but
are modified slightly in the current analysis so that imputation to the
mean will always have a score of 0. This modification makes it easier to
compare the accuracy of each imputation strategy directly with that of
imputation to the mean.

@@

The Brier score is dependent on the rate of observed events, which can
make it a difficult metric to interpret. It is often more informative to
scale the Brier score based on the Brier score of a naive model. More
specifically, for a given risk prediction model, the scaled Brier score
is computed as
\[\textrm{Scaled } \widehat{\textrm{BS}}(t)\textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t)\textrm{ of model}}{\widehat{\textrm{BS}}(t)\textrm{ of naive model}}.\]
As the Brier score for risk prediction is analogous to mean-squared
error for prediction of a continuous outcome, the scaled Brier score is
analogous to the \(R^2\) statistic. Similar to the \(R^2\) statistic, a
scaled \(\widehat{\textrm{BS}}(t)\) of 1.00 and 0.00 indicate a perfect
and worthless model, respectively. In our analyses, a Kaplan-Meier
estimate based on the training data (\textit{i.e., }a risk prediction
model that did not use any predictor variables) provided the naive
prediction. In the current analysis, we multiply scaled Brier score
values by 100 for ease of interpretation.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of 23 modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., }imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 200 replicates of Monte-Carlo
cross validation, a resampling technique for internal validation.

@@

In each replicate of Monte-Carlo cross validation, 50\% of the available
data were used for model training and testing. All predictor variables
with \textless{} 50\% missing values were considered for imputation and
subsequent model development. Among these variables, artificial
missingness (0\%, 15\%, or 30\% additional missing values) was induced
based on patient age, with younger and older patients more likely to
have missing data compared to patients who were between 40 and 65 years
of age. Prior to imputation, 50 predictor variables were selected using
a boosting model that quantified variable importance as the fractional
contribution of each predictor to the model based on the total gain
attributed to using the predictor while growing decision trees.
Imputation was conducted in the training and testing sets, separately,
for each imputation strategy. Although some imputation strategies
(\textit{e.g., }KNN and random forests) can impute data in the testing
set using models fitted to the training set, others
(\textit{e.g., }Bayesian regression, PMM, and hot deck) cannot.
Therefore, to ensure fair comparisons in our experiment, each imputation
procedure imputed data in the training set using models fitted to the
training set and then imputed data in the testing set using models
fitted to the testing set. After imputation, Cox PH and boosting models
were applied to each imputed dataset, separately. Last, model
predictions for death and transplant were computed at 6 months following
MCS surgery.

@@

SAS software (version 9.4) and Python (version 3.8.2) were used to
create analytic data for the current study
\cite{van1995python, sas2013}. Base R (version 4.0.3) was used in
combination with a number of open-source R packages
(\textit{e.g., }\texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
\texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to
conduct our analysis
\cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
Our R code is available on GitHub (see
\url{https://github.com/bcjaeger/INTERMACS-missing-data})
\cite{byron_2020_4247449}. We used Cheaha, a high performance computing
cluster at University of Alabama at Birmingham, to perform the 200
Monte-Carlo cross validation runs \cite{cheaha}.

@@

During the first 6 months after receiving MCS, the number (\% of overall
population) of deaths, transplants, and loss-to-follow-up events was
1,897 (13\%), 1,067 (7.2\%), and 981 (6.7\%), respectively. The mean
(standard deviation) age of patients was 57 (13) years, 66\% of patients
identified as white and 79\% were male (Table
\ref{tbl_characteristics}).

@@

Many predictor variables exhibited similar proportions of missing values
in different outcome groups (Table \ref{tbl_missingness}). However, the
number (percent) of missing values for surgery time was an exception,
with 6,125 (41.6\%) in the overall population, 2,246 (53.4\%) among
patients who died, and 1,713 (27.2\%) among patients who were censored.
Additionally, missing values for surgery time were frequently
accompanied by missing values for CV pressure (Figure \ref{fig:upset}).

@@

Adjusting for the amount of additional missing data amputed and the
outcome variable, the posterior probability that an imputation strategy
would improve the scaled Brier score of a downstream model relative to
imputation to the mean was maximized by using multiple imputation with
random forests (probability of improvement: 1.000; Figure
\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
random forests was estimated to have the highest posterior probability
of improving the C-index in comparison to using imputation to the mean
(probability of improvement: 1.000; Figure
\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
probability of a reduction in calibration was 1 when using either single
or multiple imputation with random forests compared with imputation to
the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
imputation using random forest was estimated to be the best overall
option, there was moderate to strong evidence that MIA and each multiple
imputation strategy we applied would improve prognostic accuracy of
downstream models compared with imputation to the mean.

@@

Previous research has also established evidence in favor of applying
multiple imputation to improve the prognostic value of risk prediction
models. For example, Hassan and Atiya demonstrated superior downstream
prediction using an ensemble multiple imputation method on synthetic
data with continuous outcomes \cite{hassan2007regression}. Similarly,
Nanni et. al demonstrated superior performance in downstream prediction
when missing values were imputed using their proposed ensemble multiple
imputation method \cite{nanni2012classifier}. Notably, the authors
artificially induced missing values in these studies and the largest
real dataset that was evaluated contained less than 700 observations. An
article by Jerez et. al evaluated missing data strategies based on the
downstream task of fitting a neural network and predicting early breast
cancer relapse \cite{jerez2010missing}. The authors found that KNN
imputation led to risk prediction models with the highest discrimination
and lowest calibration error. Results from the current study are
consistent with these previous findings but also extend their results by
providing evidence from a larger source of data
(\textit{i.e., }INTERMACS) and dealing with `real-world' missing values.

@@

\clearpage
\begin{table}
\caption{Participant characteristics stratified by event status. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
\label{tbl_characteristics}

>>>>

\author{
    Byron C. Jaeger
    \thanks{Source code available at
\url{https://github.com/bcjaeger/INTERMACS-missing-data}}
   \\
    Department of Biostatistics \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
   \And
    Ryan Cantor, PhD
   \\
    Kirklin Institute for Research in Surgical Outcomes \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Venkata Sthanam, MS
   \\
    Kirklin Institute for Research in Surgical Outcomes \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Rongbing Xie, DrPH
   \\
    Department of Surgery \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    James K. Kirklin, MD
   \\
    Kirklin Institute for Research in Surgical Outcomes \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Ramaraju Rudraraju, PhD
   \\
    Kirklin Institute for Research in Surgical Outcomes \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:ramaraju@uab.edu}{\nolinkurl{ramaraju@uab.edu}}} \\
  }

@@

\begin{abstract}
\textbf{Background} Risk prediction models play an important role in
clinical decision making. When developing risk prediction models,
practitioners often impute missing values to the mean. We evaluated the
impact of applying other strategies to impute missing values on the
prognostic accuracy of downstream risk prediction models, i.e., models
fitted to the imputed data. A secondary objective was to compare the
accuracy of imputation methods based on artificially induced missing
values. To complete these objectives, we used data from the Interagency
Registry for Mechanically Assisted Circulatory Support (INTERMACS).
\newline\textbf{Methods and Results} We applied twelve imputation
strategies in combination with two different modeling strategies for
mortality and transplant risk prediction following surgery to receive
mechanical circulatory support. Model performance was evaluated using
Monte-Carlo cross validation and measured based on outcomes 6-months
following surgery using the scaled Brier score, concordance index, and
calibration error. We used Bayesian hierarchical models to compare model
performance. Multiple imputation with random forests emerged as a robust
strategy to impute missing values, increasing model concordance by 0.003
(25th, 75th percentile: 0.0008, 0.052) compared with imputation to the
mean for mortality risk prediction using a downstream proportional
hazards model. The posterior probability that single and multiple
imputation using random forests would improve concordance versus mean
imputation was 0.464 and \textgreater0.999, respectively.
\newline\textbf{Conclusion} Selecting an optimal strategy to impute
missing values and applying multiple imputation can improve the
prognostic accuracy of downstream risk prediction models.
\end{abstract}

\keywords{
    missing data
   \and
    INTERMACS
   \and
    imputation
   \and
    heart failure
   \and
    mechanical circulatory support
   \and
    risk prediction
  }

@@

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
An important step for developing risk prediction models is dealing with
missing data, which is common in a ``real world'' database that is not
designed for the rigor of data completeness found in a clinical trial.
While some missing values are related to data entry error, others may
relate to instability of a patient's circulatory status at the time of
implant. For instance, hemodynamic laboratory values are missing for
patients in whom invasive catheterizations were not performed. Patients
without invasive catheterizations may be too ill (unstable) to tolerate
the procedure. Instead, such patients may, after determination of basic
hemodynamics, proceed directly to device placement.

The primary aim of this article is to quantify how much the prognostic
value of a risk prediction model developed from the INTERMACS registry
depends on the strategy that was applied to impute missing data prior to
developing the model. A secondary aim is to measure imputation accuracy
of each strategy by introducing varying levels of artificial missing
data (\textit{i.e., } data amputation) and then imputing it. Because
imputation to the mean has been a standard method for multiple annual
summaries of the INTERMACS data, we measure the potential improvement in
prognostic value of a risk prediction model when other strategies are
applied to impute missing data \emph{instead of imputation to the mean}.
The over-arching aim is to clarify which imputation strategies are most
likely to improve risk prediction (and by extension, quality of care)
for patients who receive MCS devices. This investigation can directly
inform future analyses of INTERMACS data and provide evidence
quantifying the benefit of imputing missing data with sound methodology.

@@

Patient follow-up begins after implantation of a durable, long term MCS
device and continues while a device is in place. Registry endpoints
include death on a device, heart transplantation, or cessation of
support (for recovery and non-recovery reasons). Mortality and
transplant after MCS were the primary outcomes for the current study. As
there were only 310 cessation of support events, we did not analyze this
outcome. For mortality and transplant risk prediction models, we applied
event-specific analysis to account for competing risks. For example, in
risk prediction models for mortality, patients were censored at time of
last contact, transplant, or cessation of support, whichever occurred
first. INTERMACS collects pre-implant data on patient characteristics,
medical status, and laboratory values. INTERMACS also collects follow-up
data at regularly scheduled visits and at occurrence of adverse events
such as re-hospitalization. For the current analysis, all pre-implant
variables were considered as potential predictors.

@@

To conduct statistical inference in the context of missing data setting,
analysts often create multiple imputed datasets, replicate an analysis
in each of them, and then pool their results to obtain valid test
statistics for hypothesis testing \cite{rubin2004multiple}. Imputation
strategies that create a single imputed dataset
(\textit{e.g., } imputation to the mean) have been shown to increase
type I errors (\textit{i.e., } rejecting a true null hypothesis) for
inferential statistics by artificially reducing the variance of observed
data and ignoring the uncertainty attributed to missing values
\cite{van2018flexible}. To ensure valid inference, imputation models
should leverage outcome variables to predict missing predictors
\cite{sterne2009multiple}. When very few data are missing, analysts may
apply listwise deletion, \textit{i.e., } removing any observation with
at least one missing value. However, listwise deletion can easily lead
to biased inference \cite{van2020rebutting}.

@@

In the presence of missing data, the goal of supervised statistical
learning is to develop a prediction function for an outcome variable
that accurately generalizes to testing data, which may or may not
contain missing values \cite{hastie2009elements}. Because testing data
may contain missing values, listwise deletion is not a feasible strategy
for statistical learning tasks. In contrast to statistical inference,
strategies that create a single imputed dataset are often used for
statistical learning \cite{kuhn2019feature}. Previous work has
emphasized the importance of imputation strategies with greater
accuracy, suggesting that more accurate imputation strategies lead to
better performance of downstream models (\textit{i.e., } models fitted
to the imputed data) \cite{jerez2010missing}. Others have shown that
imputation to the mean, model-based imputation, and multiple imputation
can provide Bayes-optimal prediction models provided certain assumptions
are true, but these assumptions are difficult to validate in applied
settings \cite{josse2019consistency}.

@@

KNN imputation identifies \(k\) `similar' observations
(\textit{i.e., } neighbors) for each observation with a missing value in
a given variable \cite{chen2000nearest}. A donor value for the current
missing value is generated by sampling or aggregating the values from
the \(k\) nearest neighbors. In the current analysis, we identified 10
nearest neighbors using Gower's distance \cite{gower}. When imputing a
single dataset using KNN, we aggregated values from 10 nearest neighbors
using the median for numeric variables and the mode for categorical
variables. When imputing multiple datasets using KNN, we aggregated
values from 2, 6, 11, 16, and 20 neighbors, separately, to create 5
imputed datasets.

@@

Random forests grow an ensemble of de-correlated decision trees, where
each tree is grown using a bootstrapped replicate of the original
training data
\cite{Breiman2001, hothorn2006survival, strobl2007bias, strobl2008conditional, ishwaran2008random, jaeger2019oblique}.
A particularly helpful feature of random forests is their ability to
estimate testing error by aggregating each decision tree's prediction
error on data outside of their bootstrapped sample
(\textit{i.e., } out-of-bag error). In the current analysis, we conduct
MICE using one random forest to impute each variable, separately. For
each imputed dataset, we allowed random forests to be re-fitted until
out-of-bag error stabilized or a maximum number of iterations was
completed. When imputing a single dataset, we used 250 trees per forest
and a maximum of 10 iterations. When imputing multiple datasets, we used
50 trees per forest and applied PMM using the random forest's predicted
values to impute missing data by sampling one value from a pool of 10
potential donors.

@@

MIA is a technique that uses missing status as a predictor rather than
explicitly imputing missing values
\cite{twala2009empirical, ding2010investigation}. MIA adds another
category to nominal variables: ``Missing.'' For numeric variables, MIA
creates two columns: one where missing values are imputed with positive
infinity and the other with negative infinity. Since PH models are not
compatible with infinite values, we only use MIA in boosting models.
When a decision tree uses a finite cut-point to split a given numeric
variable, it will assess the cut-point using both the positive and
negative infinite imputed columns, and utilize whichever column provides
the best split of the current data. This procedure translates to sending
all missing values to the left or to the right when forming two new
nodes of the decision tree, using whichever direction results in a
better split.

@@

Imputation accuracy was computed for each numerical and nominal
variable, separately. The observed values of data that were amputed were
used to assess imputation accuracy. As a consequence, we only measured
imputation accuracy 15\% or 30\% of additional missing values were
amputed. In the context of this article, the term `amputation of data'
means artificially making observed values missing. Numeric variable
imputation accuracy was measured using a re-scaled mean-squared error:
\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
MSEs are equal, and less than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
imputation accuracy was measured using a re-scaled classification
accuracy:
\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if the classification error of the current
imputation method is less than that of imputation to the mean, equal to
0 if the two classification errors are equal, and less than 0 of the
classification error of the current imputation method is worse than
imputation to the mean. These numeric and nominal scores are analogous
to the more well known \(R^2\) and Kappa statistics, respectively, but
are modified slightly in the current analysis so that imputation to the
mean will always have a score of 0. This modification makes it easier to
compare the accuracy of each imputation strategy directly with that of
imputation to the mean.

@@

The Brier score is dependent on the rate of observed events, which can
make it a difficult metric to interpret. It is often more informative to
scale the Brier score based on the Brier score of a naive model. More
specifically, for a given risk prediction model, the scaled Brier score
is computed as
\[\textrm{Scaled } \widehat{\textrm{BS}}(t)\textrm{ of model } = 1 - \frac{\widehat{\textrm{BS}}(t)\textrm{ of model}}{\widehat{\textrm{BS}}(t)\textrm{ of naive model}}.\]
As the Brier score for risk prediction is analogous to mean-squared
error for prediction of a continuous outcome, the scaled Brier score is
analogous to the \(R^2\) statistic. Similar to the \(R^2\) statistic, a
scaled \(\widehat{\textrm{BS}}(t)\) of 1.00 and 0.00 indicate a perfect
and worthless model, respectively. In our analyses, a Kaplan-Meier
estimate based on the training data (\textit{i.e., } a risk prediction
model that did not use any predictor variables) provided the naive
prediction. In the current analysis, we multiply scaled Brier score
values by 100 for ease of interpretation.

@@

To assess the prognostic value of each missing data strategy, we
internally validated a total of 23 modeling algorithm based on
combinations of imputation strategies and modeling strategies described
in Sections \ref{subsec:imputation} and \ref{subsec:modeling}. For
convenience, we use the term `modeling algorithm' to denote the
combination of a missing data strategy and a modeling strategy
(\textit{e.g., } imputation using mean/mode values followed by fitting
the PH model with stepwise variable selection) \cite{kuhn2013applied}.
We conducted internal validation using 200 replicates of Monte-Carlo
cross validation, a resampling technique for internal validation.

@@

In each replicate of Monte-Carlo cross validation, 50\% of the available
data were used for model training and testing. All predictor variables
with \textless{} 50\% missing values were considered for imputation and
subsequent model development. Among these variables, artificial
missingness (0\%, 15\%, or 30\% additional missing values) was induced
based on patient age, with younger and older patients more likely to
have missing data compared to patients who were between 40 and 65 years
of age. Prior to imputation, 50 predictor variables were selected using
a boosting model that quantified variable importance as the fractional
contribution of each predictor to the model based on the total gain
attributed to using the predictor while growing decision trees.
Imputation was conducted in the training and testing sets, separately,
for each imputation strategy. Although some imputation strategies
(\textit{e.g., } KNN and random forests) can impute data in the testing
set using models fitted to the training set, others
(\textit{e.g., } Bayesian regression, PMM, and hot deck) cannot.
Therefore, to ensure fair comparisons in our experiment, each imputation
procedure imputed data in the training set using models fitted to the
training set and then imputed data in the testing set using models
fitted to the testing set. After imputation, Cox PH and boosting models
were applied to each imputed dataset, separately. Last, model
predictions for death and transplant were computed at 6 months following
MCS surgery.

@@

SAS software (version 9.4) and Python (version 3.8.2) were used to
create analytic data for the current study
\cite{van1995python, sas2013}. These analyses were completed using The
American Heart Association Precision Medicine Platform
(\url{https://precision.heart.org/}). Base R (version 4.0.3) was used in
combination with a number of open-source R packages
(\textit{e.g., } \texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
\texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to
conduct statistical analyses and create the current manuscript
\cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
Our R code is available on GitHub (see
\url{https://github.com/bcjaeger/INTERMACS-missing-data})
\cite{byron_2020_4247449}. We used Cheaha, a high performance computing
cluster at University of Alabama at Birmingham, to perform the 200
Monte-Carlo cross validation runs \cite{cheaha}.

@@

During the first 6 months after receiving MCS, 1,897 (13\%) patients
died and 1,067 (7.2\%) had a transplant. In total, 981 (6.7\%) patients
had a follow-up time \textless{} 6 months and were considered
right-censored in the current analysis. The mean (standard deviation)
age of patients was 57 (13) years, 66\% of patients identified as white
and 79\% were male (Table \ref{tbl_characteristics}).

@@

Many predictor variables exhibited similar proportions of missing values
in different outcome groups (Table \ref{tbl_missingness}). However, the
number (percent) of missing values for surgery time was an exception,
with 6,125 (41.6\%) in the overall population, 795 (41.9\%) among
patients who died, and 110 (11.2\%) among patients who were censored.
This pattern is likely attributable to the later introduction of surgery
time into the INTERMACS registry compared to other variables. The
collection of surgery time in the INTERMACS registry began on May 30,
2014. Additionally, missing values for surgery time were frequently
accompanied by missing values for CV pressure (Figure \ref{fig:upset}).

@@

Adjusting for the amount of additional missing data amputed and the
outcome variable, the posterior probability that an imputation strategy
would improve the scaled Brier score of a downstream model relative to
imputation to the mean was maximized by using multiple imputation with
random forests (probability of improvement: \textgreater0.999; Figure
\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
random forests was estimated to have the highest posterior probability
of improving the C-index in comparison to using imputation to the mean
(probability of improvement: \textgreater0.999; Figure
\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
probability of a reduction in calibration was \textgreater0.999 when
using either single or multiple imputation with random forests compared
with imputation to the mean (Figure
\ref{fig:fig_md_strat_infer_cal_error}). Although imputation using
random forest was estimated to be the best overall option, there was
moderate to strong evidence that MIA and each multiple imputation
strategy we applied would improve prognostic accuracy of downstream
models compared with imputation to the mean.

@@

Previous research has also established evidence in favor of applying
multiple imputation to improve the prognostic value of risk prediction
models. For example, Hassan and Atiya demonstrated superior downstream
prediction using an ensemble multiple imputation method on synthetic
data with continuous outcomes \cite{hassan2007regression}. Similarly,
Nanni et. al demonstrated superior performance in downstream prediction
when missing values were imputed using their proposed ensemble multiple
imputation method \cite{nanni2012classifier}. Notably, the authors
artificially induced missing values in these studies and the largest
real dataset that was evaluated contained less than 700 observations. An
article by Jerez et. al evaluated missing data strategies based on the
downstream task of fitting a neural network and predicting early breast
cancer relapse \cite{jerez2010missing}. The authors found that KNN
imputation led to risk prediction models with the highest discrimination
and lowest calibration error. Results from the current study are
consistent with these previous findings but also extend their results by
providing evidence from a larger source of data
(\textit{i.e., } INTERMACS) and dealing with `real-world' missing
values.

@@

\paragraph{Sources of Funding}

This work is supported by the American Heart Association, grant\#
18AIML34280052. De-identified data provided from the INTERMACS contract
with from the National Heart, Lung, and Blood Institute; National
Institutes of Health; and Department of Health and Human Services, under
contract\# HHSN268201100025C.

\clearpage
\begin{table}
\footnotesize
\caption{Participant characteristics overall and stratified by event status at 6 months following surgery to receive mechanical circulatory support. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
\label{tbl_characteristics}
